Symbolic linearization of equations of motion of constrained multibody systems

### 2014-08-15 response_to_reviewers.txt

To all reviewers:
The manuscript has been drastically revised and polished. In particular, the
order of presentation and the motivation for the work has been expanded upon
greatly. Additionally, the example has been expanded upon significantly and the
steps shown more explicitly.

The electronic supplementary material should now be available and should be
executable by installing Python and SymPy (both free and open source, no other
dependencies). The simplest way to do this is to install one of the many
free scientific Python distributions which include both of these (and a number
of other useful tools). We recommend any of the following:
1) Enthought's Canopy: https://www.enthought.com/products/canopy/
2) Python XY: https://code.google.com/p/pythonxy/
3) Anacond Python: https://store.continuum.io/cshop/anaconda/
All are open source, well documented, and have free versions (and paid ones as
well with support), and have a userbase in the 10's-100's of thousands so
finding help online is generally quite easy.

To reviewer #1:
The new organization of the paper should address the concern you raised about
the presentation of linearized equations of motion without any deeper
motivation or introduction to the terms used.

Regarding the use of both independent states and dependent states, we address
this more in the introduction, the example, and the discussion of the revised
manuscript. Briefly, we formulate the equations which include both independent
and dependent states to allow for the possibility that which states are chosen
to be dependent/independent can be changed without having to redo the whole
symbolic formulation.

To reviewer #3:
Thank you for the references. We have since read all of them and highlighted in
the introduction the distinction between those works and our own.
Interestingly, we contacted Arend Schwab during the preparation of our
manuscript regarding references to linearization of constrained systems; his
response helpful, but did not include his own 2003 paper which you reference.
That aside, none of references you list employ Kane's method, nor do any
include discussion of allowing for non-trivial generalized speeds to be
selected (i.e., something other than d/dt(q_i) = u_i), nor are any of the
software packages to which the various authors refer open source -- with the
exception of the work by Minaker. The work by Minaker uses Lagrange's method,
and hence we feel is significantly distinct from our own; further he gives no
reference to the software which he refers to in his paper (Section 6),
presumably it is the software "EoM": http://137.207.14.230/vdc/index.html, but
we cannot be sure. If this is the software, it has no mailing list, uses no
source version control with a history of commits and contributions, and has no
automated test suite to verify it's correctness; it is essentially the project
of a single lab group and has no mechanism for participation by a wider
community. In contrast, the software we use to derive the equations of motion,
SymPy (sympy.org), is open source, has an extremely active mailing with
contributors from all over the world, and has been funded (through paying
students a small summer stipend) by Google for the last 6 years under the
Google Summer of Code program; it also has automated software tests for unit
tests a number example systems including the rolling disk.

Regarding the choice of independent speeds, the motivation for allowing the
choice of dependent speeds to be changed in the linearization procedure (even
though the dynamic equations, as you astutely observed, must be formed with
respect to a particular choice of dependent speeds) is so that the choice of
dependent speeds can be changed at the time of numerical evaluation rather than
at the time of symbolic derivation. We expand upon this more in the last
paragraph of the discussion, as well as provide an example of when it is useful
(computation of Lyapunov characteristic exponents for a system which a single
choice of dependent coordinates/speeds may be infeasible). Essentially, rather
than symbolically outputting the constrained equations of motion for every
possible choice of dependent states, it is possible to only output the
unconstrained equations of motion and the constraint Jacobian matrices, and
compute the "best" choice of dependent states (at runtime) and then
appropriately form the constrained dynamic equations (again, at runtime).

### 2014-08-15 response_to_reviewers_rev2.txt

Response to Reviewer #1 comments:

1. Change in revised abstract.

2. Changed in revised abstract to read: "(i.e. no DAE’s, dependent variables
are not substituted away)". We simply mean that algebraic elimination has not
been used to eliminate dependent variables.

3. We have removed the word "familiar".

4. We erased the duplicated word "system". The constraint equations are
involved as the reviewer mentions.

5. We replaced the word "types" with "levels".  Clarified
holonomic/nonholonomic and the use of time differentiated constraints at the
velocity and acceleration level.

6. We have clarified the text here.  Lagrange's method can be used for
nonoholonomic systems.  See Nonholonomic Mechanics and Control by Anthony Bloch
for extensive discussion of Lagrange's method applied to nonholonomic systems.
The two approaches (Kane's vs. Lagrange) are comparable but Kane's method
offers more flexibility on the choice of generalized speeds (i.e., u_i != dq_i
/ dt). The quoted sentence has been rewritten for clarity.

7. We have changed this equation for clarity.

8. Quoted text rewritten for clarity.  n and o are not required to be equal in
our approach, though this is commonly the case for simple holonomic systems.

9. ...

10. We have added a figure.

11.  The text has been clarified.

12.  We have revised this section to fix the reviewers comments and make the
section more clear. We do not consider the implementation/performance of a
symbolic matrix inversion.  The Jacobian of the right hand side of (44) can be
performed symbolically or numerically.  In this example, we obtained (44) by
symbolically separating and inverting terms in (43).  We symbolically obtained
(45) by computing the Jacobian using SymPy (www.sympy.org). The "steady rolling
condition" is exactly what it sounds like: the disk is rolling at a constant
spin rate, either in a straight line or along a circular path of constant
radius. See reference [14] by Oliver O'Reilly for a more detailed treatment of
rolling disks.

13. ...

14. ...

Response to Reviewer #3

1. The notation has been fixed in these equations.

2. The input description was clarified in the text following the table.

3. We have clarified this in the revised manuscript in the second paragraph
after Table 1.

4. Figure added.

5. Unnecessary step removed.

6. We have improved the discussion in the text. Having any incorrect
eigenvalues will result in an incorrect linear system response.  We do not
explore these incorrect results due to space limitations.

7. Thank you for catching this. The equation has been corrected.

8. This discrepancy has been clarified.

9. Thank you for catching this.  We added a definition for matrix M.

### 2014-08-16 PetersonGedeHubbard2012.tex
Fixed references, defined c_i, s_i
Implement changes based on reviewer 1 comments

\documentclass[smallcondensed,final]{svjour3}                     % onecolumn (standard format)
%\documentclass{article}
\RequirePackage{fix-cm}

\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage{bm}
%\usepackage{pgf}
%\usepackage[polutonikogreek]{babel}

@@

\begin{abstract}
% Context
Many common dynamic mechanical systems are subject to configuration or velocity
constraints.
% Need  (what we have and what we want)
Analyses of such systems often require linearized forms of the motion
equations.
% Task
To address this issue, we developed a procedure for organizing the constraint
and motion equations and their subsequent linearization.  This procedure was
developed specifically for equations of motion generated by Kane's method, but
is compatible with any method as long as only ordinary differential equations
are required to describe the system, and independent and dependent states
remain in the final equations (i.e. no DAE's, dependent variables are not
substituted away).
% Object of the document
Following a brief review of Kane's method and the structure of the equations it
generates, we present the procedure to symbolically linearize the nonlinear
motion equations of constrained multibody systems, and illustrate it with an
example of the rolling disk.

@@

\section{Introduction}
\label{sec:intro}
Design of linear state feedback controllers and first order stability analysis
of equilibria are two of the most common uses of multibody system dynamics
models. Both require linearized equations of motion. For unconstrained systems,
obtaining the linearized equations of motion is as simple as arranging the
equations of motion in first order form ($\dot{x} = f(x, t)$) and calculating
the Jacobian of the right hand side function, with respect to the states,
evaluated at the equilibria point of interest.  However, when the system in
questions has constraints, i.e., not all the states are independent, this
process is not as straightforward.

For a system with constraints, Kane's dynamical equations\cite{Kane1985} are
first order ODE's (in the generalized speeds) which include both the
independent and dependent states (assuming the dependent states have not been
algebraically eliminated). The kinematical and dynamical differential equations
can be solved for the time derivatives of the generalized coordinates and
generalized speeds, respectively. Taken together, it is tempting to think that
this system of first order differential equations could be linearized in the
same fashion as an unconstrained system.  Unfortunately, this is not the case.
Linearizing these equations in the same manner as one would for an
unconstrained system (by directly computing the Jacobian) is incorrect and will
yield incorrect linearized equations (and subsequently incorrect eigenvalues,
in the case of a stability analysis). We illustrate this by means of a familiar
example, the rolling disk, formulated with a non-traditional set of coordinates
and generalized speeds which are not all independent. We then provide a
systematic method to correctly generate the linear equations of motion for
constrained systems which have been found using Kane's method (or any other
method that results in equations with the same form). We apply this method to
the rolling disk example to illustrate that it yields eigenvalues that match
previously published results.

We consider three levels of constraints: configuration, velocity, and
acceleration. Configuration (holonomic) constraints limit the location or
orientation of parts of the system, relative to the external world or other
parts of the system. Velocity (nonholonomic or time differentiated holonomic)
constraints limit the speeds at which the configuration can change.
Velocity constraints often appear where there is rolling without slip or there
are closed kinematic loops. For this procedure, acceleration level constraints
are assumed to be time differentiated velocity constraints.

@@

% TODO:reiterate limitations of model
This paper presents linear, first order differential equations relating the
time derivatives of \textit{all} speeds and coordinates to the independent
speeds and coordinates, for any point of linearization.  A provided example
demonstrates the need for these equations.  We demonstrate how a reduced
set of these linearized equations can be used to perform standard stability
analysis (i.e., for control system design or linear stability analysis).  The
formalism presented here is generic enough to cover most examples of
time-varying constrained multibody systems with arbitrary external inputs and
arbitrary specified quantities.
% TODO: electronic supplementary material with examples for other methods
% TODO: rewrite following sentences after e.s.m.
The procedure was created for systems whose equations of motion have been
derived with Kane's method. Although also applicable to dynamic system
equations formulated using other methods, it is restricted to systems which can
be completely described by a set of ODE's (i.e., DAE's needn't be solved).

@@

By assumption, the velocity constraints are linear with respect to
$\mathbf{u}$, so we introduce
\begin{align}
\label{eq:constraint_B}
\mathbf{B} &\triangleq \nabla_{\mathbf{u}} \mathbf{f}_v (\mathbf{q}, \mathbf{u},
t)
\end{align}
which implies
\begin{align}
\label{eq:constraint_Bu0}
\mathbf{B}\mathbf{u} + \mathbf{f}_{vt}(\mathbf{q}, t) &= \mathbf{0}
\end{align}
where $\mathbf{f}_{vt}$ arises from explicit time varying terms that occur in
the velocity constraints. These typically arise from specified quantities whose
dynamics are being ignored (e.g., a motor spinning at constant speed).

@@

If there are velocity constraints, we must used Kane's nonholonomic dynamical
equations
%which allows Kane's nonholonomic dynamical equations to be written as
\begin{align}
\label{eq:kanes_eq_nonholonomic}
\tilde{\mathbf{F}} + \tilde{\mathbf{F}}^* = \mathbf{0}
\end{align}
where $\tilde{\mathbf{F}}$ and $\tilde{\mathbf{F}}^*$ are the nonholonomic
generalized active forces and noholonomic generalized inertia forces.  Equation
(\ref{eq:kanes_eq_nonholonomic}) will be of dimension $o - m$, where $o$ is the
total number of generalized speeds and $m$ is the number of velocity
constraints.

@@

Again to demonstrate how velocity constraints must be accounted for when
obtaining linearized equations of motion, we define the velocity of $C^*$ in
$N$ with a non-minimal choice of speeds, defined in the body-fixed ($C$) frame.
The angular velocity of $C$ in $N$, however, is expressed in the intermediate
lean frame $B$.
\begin{align}
  \label{v_u}
  {^N}\bar{v}^{C^*} &= u_4 \hat{c}_x + u_5 \hat{c}_y + u_6 \hat{c}_z \\
  \label{w_u}
  {^N}\bar{\omega}^C &= u_1 \hat{b}_x + u_2 \hat{b}_y + u_3 \hat{b}_z
\end{align}
The unconstrained partial angular velocity and partial velocity of $C^*$,
respectively, are
\begin{align}
  {^N}\bar{v}^{C^*}_u &= [0, 0, 0, \hat{c}_x, \hat{c}_y, \hat{c}_z] \\
  {^N}\bar{\omega}^C_u &= [\hat{b}_x, \hat{b}_y, \hat{b}_z, 0, 0, 0]
\end{align}
where we present the partial velocities as a $1\times6$ matrix whose entries
are populated with standard 3-vectors. Given these definitions, the velocity of
the lowest point of the disk $P^*$ is
\begin{align*}
    {^N}\bm{v}^{P^*} &= \bm{v}^{C^*} + \bm{\omega}^{C} \times r^{P^*/C^*} \\
                     &= u_4\hat{c}_x + u_5\hat{c}_y + u_6\hat{c}_z + r u_2
                        \hat{b}_x - r u_1 \hat{b}_y \\
                     &= (r u_2 c_3 + u_4) \hat{c}_x + (-r
                        u_1 + u_5)\hat{c}_y + r u_2 s_3 +
                        u6) \hat{c}_z
\end{align*}
where we have made use of the fact that $\bm{r}^{P^*/C^*} = r\bm{b}_z$.
Under the assumption of pure rolling this immediately yields three velocity
constraint equations which make up $\mathbf{f}_v = \mathbf{0}$
\begin{subequations}
\label{rd:f_v}
\begin{align}
    r u_2 c_3 + u_4 &= 0\\
            -r u_1 + u_5 &= 0\\
    r u_2 s_3 + u_6 &= 0
\end{align}
\end{subequations}
These equations are linear in the $u_i$ terms, nonlinear in the $q_i$ terms,
involve geometric system parameters (but not mass or inertial parameters), and
do not explicitly involve $\dot{q}_i$ terms. This structure will be taken
advantage of in Section \ref{sec:derivations}.

At this point, the reader familiar with the rolling disk problem is certainly
wondering why we have chosen such a diabolical set of coordinates and speeds.
In this simple example, there is an {\bf \textit{obvious and minimal}} choice of
coordinates and speeds ($q_6, u_4, u_5, u_6$, should be dependent; this choice
will be valid for all parameters and configurations; further, the derivation of
the equations of motion can be done without ever introducing these quantities),
for some systems this is not the case (e.g., Stewart's platform, bicycle). The
numeric values of system parameters and the configurations in which the system
will operate will determine which coordinates and speeds can be taken to be
dependent (and hence potentially eliminated by algebraic manipulations).
However, it may not be possible \textit{apriori} to determine easily which
coordinates and speeds should be taken as dependent; further, some systems may
operate in distinct enough regions of the configuration space that one
\textit{must} use different choices of dependent coordinates and/or speeds as
the system moves from one region to another. The purpose of our complex
choice of coordinates and speeds is to illustrate, in the context of a familiar
example, how constraints must be taken into account when linearizing equations
of motion. It is our hope that by doing this for a well known and relatively
simple example, readers can apply the same techniques to more complicated
systems where its use is more appropriate or necessary.

@@

The twelve unknown quantities appearing in the nine kinematic and dynamic
differential equations are $\dot{q}_{1-6}$ and $\dot{u}_{1-6}$. The three
acceleration constraint equations provide the final three equations which allow
for all twelve unkowns to be solved. Combining (\ref{rd:f_0_f_1}),
(\ref{rd:f_2_f_3}), and (\ref{rd:f_a}), the complete set of motion equations
can be written.
\begin{align}
\label{rd:ode}
\begin{bmatrix} \mathbf{f}_0 + \mathbf{f}_1\\
                \mathbf{f}_2 + \mathbf{f}_3\\
                \mathbf{f}_a \end{bmatrix} = \mathbf{0}
\end{align}

@@

\section{Discussion}
\label{sec:discussion}
Most modern linear control techniques assume a system can be written as
\begin{align}
\dot{\mathbf{x}} = \mathbf{A}\mathbf{x} + \mathbf{B}\mathbf{u}
\end{align}
where $\mathbf{x}$ are the system states and $\mathbf{u}$ are the control
inputs (the $\mathbf{u}$ used in control literature is analagous to the
$\mathbf{r}$ we use in this paper).  To transform equation
(\ref{eq:state_space_constrained}) into this standard form, we define
\begin{align}
\tilde{M} &=
\left[
\begin{array}{cc}
  \tilde{M}_{qq} & \bm{0}_{n \times o} \\
  \tilde{M}_{uqc} & \tilde{M}_{uuc} \\
  \tilde{M}_{uqd} & \tilde{M}_{uud}
\end{array}
\right]\\
\tilde{A} &=
   \left[
     \begin{array}{cc}
       (\tilde{A}_{qq} + \tilde{A}_{qu} C_1 ) C_0 & \tilde{A}_{qu} C_2 \\
       (\tilde{A}_{uqc} + \tilde{A}_{uuc} C_1 ) C_0 & \tilde{A}_{uuc} C_2\\
       (\tilde{A}_{uqd} + \tilde{A}_{uud} C_1 ) C_0 & \tilde{A}_{uud} C_2
     \end{array}
   \right]\\
\tilde{B} &= 
    \left[
      \begin{array}{c}
        \bm{0}_{(n+m) \times s} \\
        \tilde{B}_{u}
      \end{array}
    \right]
\end{align}
from (\ref{eq:state_space_constrained}) and
\begin{align}
  \label{eq:A_prime}
    A^\prime &\triangleq \tilde{M}^{-1} \tilde{A} \\
  \label{eq:B_prime}
    B^\prime &\triangleq \tilde{M}^{-1} \tilde{B}
\end{align}
where  $A^\prime \in \bm{R}^{(o + n) \times (o - m + n -l)}$, $B^\prime \in
\bm{R}^{(o + n) \times s}$.  We can extract the rows corresponding to the
independent states by defining
\begin{align}
  \label{eq:P_prime}
    P^\prime &\triangleq \begin{bmatrix}
        P_{qi} & \bm{O}_{n \times (o - m)} \\
        \bm{O}_{o \times (n - l)} & P_{ui}
    \end{bmatrix} \\
  \label{eq:A}
    A &\triangleq P^{\prime T} A^\prime \\
  \label{eq:B}
    B &\triangleq P^{\prime T} B^\prime
\end{align}
where $P^\prime \in \bm{R}^{(o + n) \times (o - m + n - l)}$.  Defining
$\bm{x}_i = \left[\delta\bm{q}_i,\,\delta\bm{u}_i\right]^{T}$ yields the square
state space system $\dot{\bm{x}}_i = A \bm{x}_i + B \bm{r}$ to which standard
linear systems analyses may applied.  It is worth noting that the rows of
$A^\prime$ and $B^\prime$ which correspond to dependent states can be used in
output or measurement equations of a linear state space model (e.g.,
measurements from an accelerometer, or measurements of dependent speeds).

@@

A task where the ability to switch the choice of dependent states ``on the
fly'' is extremely helpful is in the computation of Lyapunov characteristic
exponents (LCE's). When computing LCE's, time simulation of the nonlinear
dynamics \textit{and} linearization at each point along the trajectory a
required~\cite{Benettin1980a,Benettin1980b,Udwadia2001}; having a computer code
that makes it easy to switch choice of dependent states at the time of
numerical evaluation is of great benefit in this case.

>>>>

\documentclass[smallcondensed,final]{svjour3}                     % onecolumn (standard format)
\RequirePackage{fix-cm}

\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage{bm}

@@

\begin{abstract}
% Context
Many common dynamic mechanical systems are subject to configuration or velocity
constraints.
% Need  (what we have and what we want)
Analyses of such systems often require linearized forms of the motion
equations.
% Task
To address this issue, we developed a procedure for organizing the constraint
and motion equations and their subsequent linearization.  This procedure was
developed for equations of motion generated by Kane's method where dependent
states have not been algebraically eliminated; it is compatible with any method
as long as only ordinary differential equations are required to describe the
system (i.e., no differential algebraic equations).
% Object of the document
Following a brief review of Kane's method and the structure of the equations it
generates, we present the procedure to symbolically linearize the nonlinear
motion equations of constrained multibody systems, and illustrate it with an
example of the rolling disk.

@@

\section{Introduction}
\label{sec:intro}
Design of linear state feedback controllers and first order stability analysis
are common uses of multibody system dynamics models. Both require linearized
equations of motion. For unconstrained systems, obtaining the linearized
equations of motion is as simple as arranging them  in first order form
($\dot{x} = f(x, t)$) and evaluating $\frac{\partial f}{\partial x}}$ at the
point of interest.  However, when the system has constraints not all the states
are independent, and this process is not as straightforward.

For a system with constraints, Kane's dynamical equations\cite{Kane1985} are
first order ODEs (in the generalized speeds) which include both the
independent and dependent states (assuming the dependent states have not been
algebraically eliminated). The kinematical and dynamical differential equations
can be solved for the time derivatives of the generalized coordinates and
generalized speeds, respectively. Taken together, it is tempting to think that
this system of first order differential equations could be linearized in the
same fashion as an unconstrained system.  Unfortunately, this is not the case.
Linearizing these equations in the same manner as one would for an
unconstrained system (by directly computing the Jacobian) is incorrect and will
yield incorrect linearized equations (and subsequently incorrect eigenvalues,
in the case of a stability analysis). We illustrate this by means of a familiar
example, the rolling disk, formulated with a non-traditional set of coordinates
and generalized speeds which are not all independent. We then provide a
systematic method to correctly generate the linear equations of motion for
constrained systems which have been found using Kane's method (or any other
method that results in equations with the same form). We apply this method to
the rolling disk example to illustrate that it yields eigenvalues that match
previously published results.

We consider three levels of constraints: configuration, velocity, and
acceleration. Constraints at the configuration level (holonomic) limit the
location or orientation of parts of the system. Configuration constraints may
arise from closed kinematic loops. Constraints at the velocity level limit
permissible velocities of the system. Velocity constraints may arise from
rolling without slip (nonholonomic) or closed kinematic loops (time
differentiated holonomic).  Constraints at the acceleration level are assumed
to be time differentiated velocity level constraints.

@@

% TODO:reiterate limitations of model
This paper presents linear, first order differential equations relating the
time derivatives of \textit{all} speeds and coordinates to the independent
speeds and coordinates, for any point of linearization.  A provided example
demonstrates the need for these equations.  We demonstrate how a reduced
set of these linearized equations can be used to perform standard stability
analysis (i.e., for control system design or linear stability analysis).  The
formalism presented here is generic enough to cover most examples of
time-varying constrained multibody systems with arbitrary external inputs and
arbitrary specified quantities.
% TODO: electronic supplementary material with examples for other methods
% TODO: rewrite following sentences after e.s.m.
The procedure was created for systems whose equations of motion have been
derived with Kane's method. Although also applicable to dynamic system
equations formulated using other methods, it is restricted to systems which can
be completely described by a set of ODEs (i.e., DAEs needn't be solved).

@@

The velocity constraints are assumed linear with respect to
$\mathbf{u}$, so we introduce
\begin{align}
\label{eq:constraint_B}
\mathbf{B} &\triangleq \nabla_{\mathbf{u}} \mathbf{f}_v (\mathbf{q}, \mathbf{u},
t)
\end{align}
where $\mathbf{B}$ is implicitly dependent upon $\mathbf{q}$ and $t$.  The velocity
constraints can then be expressed as
\begin{align}
\label{eq:constraint_Bu0}
\mathbf{B}\mathbf{u} + \mathbf{f}_{vt}(\mathbf{q}, t) &= \mathbf{0}
\end{align}
where $\mathbf{f}_{vt}$ arises from explicit time varying terms that occur in
the velocity constraints. These typically arise from specified quantities whose
dynamics are being ignored (e.g., a motor spinning at constant speed).

@@

If there are velocity constraints, we must used Kane's nonholonomic dynamical
equations
%which allows Kane's nonholonomic dynamical equations to be written as
\begin{align}
\label{eq:kanes_eq_nonholonomic}
\tilde{\mathbf{F}} + \tilde{\mathbf{F}}^* = \mathbf{0}
\end{align}
where $\tilde{\mathbf{F}}$ and $\tilde{\mathbf{F}}^*$ are the nonholonomic
generalized active forces and nonholonomic generalized inertia forces.  Equation
(\ref{eq:kanes_eq_nonholonomic}) will be of dimension $o - m$, where $o$ is the
total number of generalized speeds and $m$ is the number of velocity
constraints.

@@

Again to demonstrate how velocity constraints must be accounted for when
obtaining linearized equations of motion, we define the velocity of $C^*$ in
$N$ with a non-minimal choice of speeds, defined in the body-fixed ($C$) frame.
The angular velocity of $C$ in $N$, however, is expressed in the intermediate
lean frame $B$.
\begin{align}
  \label{v_u}
  {^N}\bar{v}^{C^*} &= u_4 \hat{c}_x + u_5 \hat{c}_y + u_6 \hat{c}_z \\
  \label{w_u}
  {^N}\bar{\omega}^C &= u_1 \hat{b}_x + u_2 \hat{b}_y + u_3 \hat{b}_z
\end{align}
The unconstrained partial angular velocity and partial velocity of $C^*$,
respectively, are
\begin{align}
  {^N}\bar{v}^{C^*}_u &= [0, 0, 0, \hat{c}_x, \hat{c}_y, \hat{c}_z] \\
  {^N}\bar{\omega}^C_u &= [\hat{b}_x, \hat{b}_y, \hat{b}_z, 0, 0, 0]
\end{align}
where we present the partial velocities as a $1\times6$ matrix whose entries
are populated with standard 3-vectors. Given these definitions, the velocity of
the lowest point of the disk $P^*$ is
\begin{align*}
    {^N}\bm{v}^{P^*} &= \bm{v}^{C^*} + \bm{\omega}^{C} \times r^{P^*/C^*} \\
                     &= u_4\hat{c}_x + u_5\hat{c}_y + u_6\hat{c}_z + r u_2
                        \hat{b}_x - r u_1 \hat{b}_y \\
                     &= (r u_2 c_3 + u_4) \hat{c}_x + (-r
                        u_1 + u_5)\hat{c}_y + (r u_2 s_3 +
                        u_6) \hat{c}_z
\end{align*}
where we have made use of the fact that $\bm{r}^{P^*/C^*} = r\bm{b}_z$.
Under the assumption of pure rolling this immediately yields three velocity
constraint equations which make up $\mathbf{f}_v = \mathbf{0}$
\begin{subequations}
\label{rd:f_v}
\begin{align}
    r u_2 c_3 + u_4 &= 0\\
            -r u_1 + u_5 &= 0\\
    r u_2 s_3 + u_6 &= 0
\end{align}
\end{subequations}
These equations are linear in the $u_i$ terms, nonlinear in the $q_i$ terms,
involve geometric system parameters (but not mass or inertial parameters), and
do not explicitly involve $\dot{q}_i$ terms. This structure will be taken
advantage of in Section \ref{sec:derivations}.

At this point, the reader familiar with the rolling disk problem may wondering
why we have purposefully chosen a non-minimal set of coordinates and speeds.
In this simple example, there is an {\bf \textit{obvious and minimal}} choice
of coordinates and speeds ($q_6, u_4, u_5, u_6$, should be dependent; this
choice will be valid for all parameters and configurations; further, the
derivation of the equations of motion can be done without ever introducing
these quantities), for some systems this is not the case (e.g., Stewart's
platform, bicycle). The numeric values of system parameters and the
configurations in which the system will operate will determine which
coordinates and speeds can be taken to be dependent (and hence potentially
eliminated by algebraic manipulations).  However, it may not be possible
\textit{a priori} to determine easily which coordinates and speeds should be
taken as dependent; further, some systems may operate in distinct enough
regions of the configuration space that one \textit{must} use different choices
of dependent coordinates and/or speeds as the system moves from one region to
another. The purpose of our complex choice of coordinates and speeds is to
illustrate, in the context of a familiar example, how constraints must be taken
into account when linearizing equations of motion. It is our hope that by doing
this for a well known and relatively simple example, readers can apply the same
techniques to more complicated systems where its use is more appropriate or
necessary.

@@

The twelve unknown quantities appearing in the nine kinematic and dynamic
differential equations are $\dot{q}_{1-6}$ and $\dot{u}_{1-6}$. The three
acceleration constraint equations provide the final three equations which allow
for all twelve unknowns to be solved. Combining (\ref{rd:f_0_f_1}),
(\ref{rd:f_2_f_3}), and (\ref{rd:f_a}), the complete set of motion equations
can be written.
\begin{align}
\label{rd:ode}
\begin{bmatrix} \mathbf{f}_0 + \mathbf{f}_1\\
                \mathbf{f}_2 + \mathbf{f}_3\\
                \mathbf{f}_a \end{bmatrix} = \mathbf{0}
\end{align}

@@

\section{Discussion}
\label{sec:discussion}
Most modern linear control techniques assume a system can be written as
\begin{align}
\dot{\mathbf{x}} = \mathbf{A}\mathbf{x} + \mathbf{B}\mathbf{u}
\end{align}
where $\mathbf{x}$ are the system states and $\mathbf{u}$ are the control
inputs (the $\mathbf{u}$ used in control literature is analogous to the
$\mathbf{r}$ we use in this paper).  To transform equation
(\ref{eq:state_space_constrained}) into this standard form, we define
\begin{align}
\tilde{M} &=
\left[
\begin{array}{cc}
  \tilde{M}_{qq} & \bm{0}_{n \times o} \\
  \tilde{M}_{uqc} & \tilde{M}_{uuc} \\
  \tilde{M}_{uqd} & \tilde{M}_{uud}
\end{array}
\right]\\
\tilde{A} &=
   \left[
     \begin{array}{cc}
       (\tilde{A}_{qq} + \tilde{A}_{qu} C_1 ) C_0 & \tilde{A}_{qu} C_2 \\
       (\tilde{A}_{uqc} + \tilde{A}_{uuc} C_1 ) C_0 & \tilde{A}_{uuc} C_2\\
       (\tilde{A}_{uqd} + \tilde{A}_{uud} C_1 ) C_0 & \tilde{A}_{uud} C_2
     \end{array}
   \right]\\
\tilde{B} &=
    \left[
      \begin{array}{c}
        \bm{0}_{(n+m) \times s} \\
        \tilde{B}_{u}
      \end{array}
    \right]
\end{align}
from (\ref{eq:state_space_constrained}) and
\begin{align}
  \label{eq:A_prime}
    A^\prime &\triangleq \tilde{M}^{-1} \tilde{A} \\
  \label{eq:B_prime}
    B^\prime &\triangleq \tilde{M}^{-1} \tilde{B}
\end{align}
where  $A^\prime \in \bm{R}^{(o + n) \times (o - m + n -l)}$, $B^\prime \in
\bm{R}^{(o + n) \times s}$.  We can extract the rows corresponding to the
independent states by defining
\begin{align}
  \label{eq:P_prime}
    P^\prime &\triangleq \begin{bmatrix}
        P_{qi} & \bm{O}_{n \times (o - m)} \\
        \bm{O}_{o \times (n - l)} & P_{ui}
    \end{bmatrix} \\
  \label{eq:A}
    A &\triangleq P^{\prime T} A^\prime \\
  \label{eq:B}
    B &\triangleq P^{\prime T} B^\prime
\end{align}
where $P^\prime \in \bm{R}^{(o + n) \times (o - m + n - l)}$.  Defining
$\bm{x}_i = \left[\delta\bm{q}_i,\,\delta\bm{u}_i\right]^{T}$ yields the square
state space system $\dot{\bm{x}}_i = A \bm{x}_i + B \bm{r}$ to which standard
linear systems analyses may applied.  It is worth noting that the rows of
$A^\prime$ and $B^\prime$ which correspond to dependent states can be used in
output or measurement equations of a linear state space model (e.g.,
measurements from an accelerometer, or measurements of dependent speeds).

@@

A task where the ability to switch the choice of dependent states ``on the
fly'' is extremely helpful is in the computation of Lyapunov characteristic
exponents (LCEs). When computing LCEs, time simulation of the nonlinear
dynamics \textit{and} linearization at each point along the trajectory a
required~\cite{Benettin1980a,Benettin1980b,Udwadia2001}; having a computer code
that makes it easy to switch choice of dependent states at the time of
numerical evaluation is of great benefit in this case.
